Metal Mixing and Ejection in Dwarf Galaxies is Dependent on Nucleosynthetic Source

### 2018-07-30 main.tex
fixing HIM

One-zone chemical evolution models assume homoegeneous mixing of metals from recent star formation into gas available for future star formation \citep[e.g.][]{Lanfranchi2006b,Kirby2011-III,Cote2017,Andrews2017}. While this incorrectly assumes that metal mixing in the ISM plays no role in delaying future enrichment in star formation, accounting for this process is challening. More complex models \citep[e.g.][]{SchonrichBinney2009,Pezzulli2016} employ multiple (usually radial) zones or follow multiple gas phases to attempt to account for these effects. How to properly account for multi-phase mixing, however, is still poorly understood. This is in large part because both these models and large-scale cosmological simulations lack the necessary fidelity to capture the detailed, multi-phase mixing process of metals in the ISM directly. Recent hydrodynamics simulations have employed parameteric models to account for unresolved sub-grid metal mixing \citep{PanScannapiecoScalo2013,Sarmento2017}, which plays an important role in determining the chemical properties of galaxies \citep[e.g.][]{Shen2010, Pilkington2012, Few2012, Brook2014, Escala2018}. Consequently, following metal mixing, if treated in these models, requires assumptions about how metals mix within the ISM and how metals couple to galactic outflows. When metals are tracked individually, as opposed to a global metallicity field, their injection, mixing, and outflow properties are often treated uniformly. However, these processes are critical components in understanding galactic chemical evolution. Detailed hydrodynamics simulations incorporating a realistic multi-phase ISM and detailed stellar feedback are required to understand metal mixing and metal outflows \citep[e.g.][]{FengKrumholz2014,Armillotta2018}. In addition, it remains to be seen to what degree, if at all, metals of different nucleosynthetic origins may couple differently to the ISM.

@@

The top panel of Figure~\ref{fig:species_fractions} gives the mass fraction of each element at the end of the simulation (500~Myr) in each of four reservoirs: locked in stars (yellow), in the ISM (blue), outside the galaxy but within the virial radius (purple), and outside the virial radius (salmon), including gas that has left the domain. We subdivide the ISM by phase in the bottom panel, giving the mass fractions of each element in the cold neutral medium (CNM, $f_{\rm H_2} < 0.5$,  T$< 100$~K), warm neutral medium (WNM, $10^2~\rm{K}\le \rm{T} < 10^4~\rm{K}$), warm ionized medium (WIM, $10^4~\rm{K}\le \rm{T} < 10^{5.5}~\rm{K}$), the hot ionized medium (H~{\sc i}M T$\ge 10^{5.5}$~K), and locked in stars (yellow). Again, with the exception of H and He, we only consider metals produced self-consistently through our star formation and stellar feedback methods; the initial mass of each metal species is zero.

@@

\begin{figure*}
\centering
\includegraphics[width=0.95\linewidth]{DD0390_element_by_element.png}
\caption{The numerical PDFs (solid histograms) and the associated log-normal + power-law fits (dashed lines) for a subset of the elements tracked in our simulation in each of the four gas phases defined in Section~\ref{sec:ejection}: CNM (blue), WNM (green), WIM (purple), H~{\sc i}M (red), and all the gas in the ISM (black). For clarity, each distribution is normalized to the mode of the full-disk PDF (black) and is centered on the median value of the full-disk PDF. We note the vertical axis normalization is such that integrating over the shown PDF gives the mass fraction of that phase in the disk. Since the CNM dominates the mass fraction of our galaxy, the black curve is often obscured at low metal fractions.}
\label{fig:log-normal}
\end{figure*}

@@

Unsurprisingly the diffuse H~{\sc i}M contains the most metal enriched gas, as it is comprised predominately of metal enriched supernova ejecta. Clearly this leads to long power-law tails towards high metal fractions for each species, with a very narrow, poorly defined peak at low metal fractions. Generally, in colder gas the metal PDFs become less enriched with broader low metal fraction components and steeper, power-law tails. Although the extended power-law tail of the H~{\sc i}M leads to a large range of metal fractions, the H~{\sc i}M represents very little mass, and this tail represents recent, un-mixed enrichment. The comparatively narrow width in the log-normal regimes of CNM is perhaps surprising. Although this gas is at lower metal fraction than the WNM and WIM, it would appear that it is more well mixed. This runs counter to the idea that mixing times should be long in colder gas unless mixing first proceeds rapidly in the hot phases before mixing in with cold, disconnected structures across the galaxy.

Although these trends across phases hold for all elements, there are qualitative differences between elements, particularly between those ejected predominantly by AGB winds (e.g. Ba and N) and those ejected by supernovae (e.g. O and Mg). In all phases, except the H~{\sc i}M, the AGB-wind elements have broader distributions that are less well described by our adopted $p(Z)$ than the metals dominated by supernova enrichment. The power-law component of N and Ba is generally shallower than all other metals across each phase (except the H~{\sc i}M), particularly in the CNM. Ba and N do not show significant differences among the rest of the metals in the H~{\sc i}M, though this is likely because the Ba and N present in this phase is dominated by the Ba and N included in supernovae yields. Again these differences between yield sources could be driven both by differences in their enrichment timescales and therefore differences in the typical ISM environment each event encounters, and the differences in energetics between AGB winds and supernovae. AGB wind elements enrich the WNM and WIM directly, rather than the H~{\sc i}M, which would increase their mixing timescales and broaden their PDFs.

@@

For O, the median mass fraction is ordered by phase temperature. The H~{\sc i}M is significantly more enriched than the cooler phases by anywhere from 0.1 dex to 4 dex, fluctuating by $\pm$ 1 dex over the simulation time. The frequent large skew in the H~{\sc i}M (see second panel) and spread in the H~{\sc i}M (right two panels), coupled with its continual fluctuation, suggests that the H~{\sc i}M is not in equilibrium.
%This phase is poorly mixed because it is short lived, either cooling into different phases of the ISM or being ejected from the galaxy, and is continually enriched by new SN events.
Each cooler phase in O is progressively less enriched (lower median), with smaller skew and spread. The offset between phases and increasingly well mixed gas from hot to cold indicates that metal enrichment in the ISM of galaxies proceeds first through mixing on large scales in a hot phase, before progressively cooling through multiple phases until enriching star forming gas. This is the only explanation for how the cold gas can rapidly homogenize over the whole galaxy within $\sim$50 Myr, roughly when the cold phase exhibits a nearly constant spread (right panels). Individual enrichment events will have significantly higher metallicities than the ambient ISM in any phase, and will drive an increase in the difference between the mean and median of the PDF. These can be seen as the obvious spikes in the H~{\sc i}M and WIM. For O, the lack of these spikes in the two cold phases suggests again that enrichment does not occur directly in these phases, but proceeds more gradually through the warmer phases first.

@@

Although these trends are generally true for Ba, its evolution is much more complicated.
%In addition, the H~{\sc i}M is much less offset from the rest of the phases than is seen in O, with much less variability.
Unlike O, the spread and skew in Ba for the CNM, WNM, and WIM \textit{increase} in during the evolution, reaching differences in 90$^{\rm th}$ and 10$^{\rm th}$ percentiles of nearly 2 dex in the CNM and over 2 dex in the WNM and WIM. The H~{\sc i}M is seemingly unaffected by this trend, and simply fluctuates throughout the simulation. Given their lower energies, AGB winds more directly enrich the WNM and WIM, not the H~{\sc i}M as in supernovae. The consequence of this is clear in Figure~\ref{fig:log-normal} by the wider PDFs and longer power-law tails in Ba in these phases. These tails represent the most recently enriched gas, and is clearly much more locally confined than O.

@@

To summarize, this figure demonstrates: 1) there are qualitative differences in how supernova injected elements (e.g. O, Mg) and AGB wind injected elements (e.g. N, Ba) are distributed through the ISM, with the latter having a broader range of variation and being less well mixed in all phases except the H~{\sc i}M, 2) hotter phases are more metal enriched, both because the cooler phases make up most of the initially unenriched mass of the ISM, and the hotter phases are more directly populated by recent enrichment events, 3) the cooler, denser phases, particularly for supernova injected elements, are more well mixed than the hot phases of the ISM, 4) the PDFs of metal mass fraction are best fit by a log-normal + power law function, and therefore 5) the median metallicity available for star formation rates lies below the mean galactic value. In the case of supernova injected elements, enrichment proceeds quickly through the H~{\sc i}M over the entire galaxy. For AGB injected elements, enrichment proceeds through the WIM and WNM, leading to longer mixing timescales and larger metal fraction variations in the ISM.

@@

This toy model provides a physical intuition for the general trends in the PDFs presented here across ISM phases. The H~{\sc i}M is comprised predominantly of gas from individual enrichment events that, due to its low density, easily create large power-law tails beginning at very high metal fractions. Whatever ambient component of the H~{\sc i}M that exists is well mixed, leading to a narrow and sub-dominant log-normal component at low-metallicities. Cold, dense gas, which is almost never directly impacted by these individual enrichment events, is enriched almost entirely by diffusion, and thus has an almost completely log-normal PDF with very little, if any, power-law tail.

@@

%
% A.E. paragraph may need work
%
Mixing timescales are likely proportional to the eddy turnover times at the injection scale \citep{PanScannapieco2010, Colbrook2017} and the properties of turbulence in the ISM \citep{YangKrumholz2012}. Mixing within a phase is likely dependent upon the phase's sound speed and velocity dispersion. This would imply rapid mixing timescales in the WIM and H~{\sc i}M, with typical velocity dispersions of $\sim$~30 km~s$^{-1}$ and $\sim$~100~km~s$^{-1}$ respectively, and long mixing timescales in cold, dense gas ($\sim$ 1 km s$^{-1}$). Our results show, however, that in general the WIM and H~{\sc i}M are the least well mixed, while the CNM is the most well mixed across the galaxy. This is particularly curious as the sound crossing time of the H~{\sc i}M across the galaxy ($\sim 1$ kpc) is $\sim$10~Myr, compared to $\sim$~1~Gyr for the coldest gas. It must be that the H~{\sc i}M is far from equillibrium throughout the simulation, in part due to the continual enrichment by ongoing supernovae. Newly enriched gas must mix through the H~{\sc i}M on galaxy scales, becoming well mixed and less enriched by the time it cools into the CNM and eventually star forming gas. Since elements produced in AGB winds do not directly enter the H~{\sc i}M, mixing is less efficient driving larger variations across the galaxy. The idea of metals processing first through the hot ISM has been proposed before to explain observed metallicity trends in the outskirts of more massive galaxies \citep{Tassis2008,Werk2011}.

@@

Previous work on the evolution of metals in the ISM varies from studies concerning the advection of passive scalars in idealized turbulent boxes \citep[e.g.][]{Pope1991, PanScannapieco2010, PanScannapiecoScalo2012, PanScannapiecoScalo2013, YangKrumholz2012, SurPanScannapieco2014, Colbrook2017}
% cite passive scalar reviews by Warhaft 2000 in fluid mech and scalo + Elmegreen 2004?
to global galaxy models studying generalized advection and mixing of passive scalars \citep[e.g.][]{deAvillez2002,Petit2015,Goldbaum2016} to models with more detailed self-consistent metal enrichment \citep[e.g.][]{Revaz2009,Escala2018}.
% summarize conclusions from these works:

@@

More detailed models of global galaxy evolution have generally focused on spatial correlations and mixing timescales of initially asymmetric fields, without concern for ongoing enrichment \citep[e.g.][]{deAvillez2002,Petit2015} or focused primarily on the evolution of stellar abundance patterns and metallicity distribution functions \citep[e.g][need more]{Jeon2017,Hirai2017,Escala2018}, without directly examining the evolution of gas-phase metallicity PDFs.

@@

%We discuss possible caveats to this in Section~\ref{sec:caveats}. Does this extend to more massive? how does DM potential affect this? Are general trends still true in much larger galaxies? or are things very different once H~{\sc i}M crossing times are more like 100 Myr (aka MW size galaxies).

>>>>

One-zone chemical evolution models assume homoegeneous mixing of metals from recent star formation into gas available for future star formation \citep[e.g.][]{Lanfranchi2006b,Kirby2011-III,Cote2017,Andrews2017}. While this assumes that metal mixing in the ISM plays no role in delaying future enrichment in star formation, accounting for this process is challening. More complex models \citep[e.g.][]{SchonrichBinney2009,Pezzulli2016} employ multiple (usually radial) zones or follow multiple gas phases to attempt to account for these effects. How to properly account for multi-phase mixing, however, is still poorly understood. This is in large part because both these models and large-scale cosmological simulations lack the necessary fidelity to capture the detailed, multi-phase mixing process of metals in the ISM directly. Recent hydrodynamics simulations have employed parameteric models to account for unresolved sub-grid metal mixing \citep{PanScannapiecoScalo2013,Sarmento2017}, which plays an important role in determining the chemical properties of galaxies \citep[e.g.][]{Shen2010, Pilkington2012, Few2012, Brook2014, Escala2018}. Consequently, following metal mixing, if treated in these models, requires assumptions about how metals mix within the ISM and how metals couple to galactic outflows. When metals are tracked individually, as opposed to a global metallicity field, their injection, mixing, and outflow properties are often treated uniformly. However, these processes are critical components in understanding galactic chemical evolution. Detailed hydrodynamics simulations incorporating a realistic multi-phase ISM and detailed stellar feedback are required to understand metal mixing and metal outflows \citep[e.g.][]{FengKrumholz2014,Armillotta2018}. In addition, it remains to be seen to what degree, if at all, metals of different nucleosynthetic origins may couple differently to the ISM.

@@

The top panel of Figure~\ref{fig:species_fractions} gives the mass fraction of each element at the end of the simulation (500~Myr) in each of four reservoirs: locked in stars (yellow), in the ISM (blue), outside the galaxy but within the virial radius (purple), and outside the virial radius (salmon), including gas that has left the domain. We subdivide the ISM by phase in the bottom panel, giving the mass fractions of each element in the cold neutral medium (CNM, $f_{\rm H_2} < 0.5$,  T$< 100$~K), warm neutral medium (WNM, $10^2~\rm{K}\le \rm{T} < 10^4~\rm{K}$), warm ionized medium (WIM, $10^4~\rm{K}\le \rm{T} < 10^{5.5}~\rm{K}$), the hot ionized medium (HIM T$\ge 10^{5.5}$~K), and locked in stars (yellow). Again, with the exception of H and He, we only consider metals produced self-consistently through our star formation and stellar feedback methods; the initial mass of each metal species is zero.

@@

\begin{figure*}
\centering
\includegraphics[width=0.95\linewidth]{DD0390_element_by_element.png}
\caption{The numerical PDFs (solid histograms) and the associated log-normal + power-law fits (dashed lines) for a subset of the elements tracked in our simulation in each of the four gas phases defined in Section~\ref{sec:ejection}: CNM (blue), WNM (green), WIM (purple), HIM (red), and all the gas in the ISM (black). For clarity, each distribution is normalized to the mode of the full-disk PDF (black) and is centered on the median value of the full-disk PDF. We note the vertical axis normalization is such that integrating over the shown PDF gives the mass fraction of that phase in the disk. Since the CNM dominates the mass fraction of our galaxy, the black curve is often obscured at low metal fractions.}
\label{fig:log-normal}
\end{figure*}

@@

Unsurprisingly the diffuse HIM contains the most metal enriched gas, as it is comprised predominately of metal enriched supernova ejecta. Clearly this leads to long power-law tails towards high metal fractions for each species, with a very narrow, poorly defined peak at low metal fractions. Generally, in colder gas the metal PDFs become less enriched with broader low metal fraction components and steeper, power-law tails. Although the extended power-law tail of the HIM leads to a large range of metal fractions, the HIM represents very little mass, and this tail represents recent, un-mixed enrichment. The comparatively narrow width in the log-normal regimes of CNM is perhaps surprising. Although this gas is at lower metal fraction than the WNM and WIM, it would appear that it is more well mixed. This runs counter to the idea that mixing times should be long in colder gas unless mixing first proceeds rapidly in the hot phases before mixing in with cold, disconnected structures across the galaxy.

Although these trends across phases hold for all elements, there are qualitative differences between elements, particularly between those ejected predominantly by AGB winds (e.g. Ba and N) and those ejected by supernovae (e.g. O and Mg). In all phases, except the HIM, the AGB-wind elements have broader distributions that are less well described by our adopted $p(Z)$ than the metals dominated by supernova enrichment. The power-law component of N and Ba is generally shallower than all other metals across each phase (except the HIM), particularly in the CNM. Ba and N do not show significant differences among the rest of the metals in the HIM, though this is likely because the Ba and N present in this phase is dominated by the Ba and N included in supernovae yields. Again these differences between yield sources could be driven both by differences in their enrichment timescales and therefore differences in the typical ISM environment each event encounters, and the differences in energetics between AGB winds and supernovae. AGB wind elements enrich the WNM and WIM directly, rather than the HIM, which would increase their mixing timescales and broaden their PDFs.

@@

For O, the median mass fraction is ordered by phase temperature. The HIM is significantly more enriched than the cooler phases by anywhere from 0.1 dex to 4 dex, fluctuating by $\pm$ 1 dex over the simulation time. The frequent large skew in the HIM (see second panel) and spread in the HIM (right two panels), coupled with its continual fluctuation, suggests that the HIM is not in equilibrium.
%This phase is poorly mixed because it is short lived, either cooling into different phases of the ISM or being ejected from the galaxy, and is continually enriched by new SN events.
Each cooler phase in O is progressively less enriched (lower median), with smaller skew and spread. The offset between phases and increasingly well mixed gas from hot to cold indicates that metal enrichment in the ISM of galaxies proceeds first through mixing on large scales in a hot phase, before progressively cooling through multiple phases until enriching star forming gas. This is the only explanation for how the cold gas can rapidly homogenize over the whole galaxy within $\sim$50 Myr, roughly when the cold phase exhibits a nearly constant spread (right panels). Individual enrichment events will have significantly higher metallicities than the ambient ISM in any phase, and will drive an increase in the difference between the mean and median of the PDF. These can be seen as the obvious spikes in the HIM and WIM. For O, the lack of these spikes in the two cold phases suggests again that enrichment does not occur directly in these phases, but proceeds more gradually through the warmer phases first.

@@

Although these trends are generally true for Ba, its evolution is much more complicated.
%In addition, the HIM is much less offset from the rest of the phases than is seen in O, with much less variability.
Unlike O, the spread and skew in Ba for the CNM, WNM, and WIM \textit{increase} in during the evolution, reaching differences in 90$^{\rm th}$ and 10$^{\rm th}$ percentiles of nearly 2 dex in the CNM and over 2 dex in the WNM and WIM. The HIM is seemingly unaffected by this trend, and simply fluctuates throughout the simulation. Given their lower energies, AGB winds more directly enrich the WNM and WIM, not the HIM as in supernovae. The consequence of this is clear in Figure~\ref{fig:log-normal} by the wider PDFs and longer power-law tails in Ba in these phases. These tails represent the most recently enriched gas, and is clearly much more locally confined than O.

@@

To summarize, this figure demonstrates: 1) there are qualitative differences in how supernova injected elements (e.g. O, Mg) and AGB wind injected elements (e.g. N, Ba) are distributed through the ISM, with the latter having a broader range of variation and being less well mixed in all phases except the HIM, 2) hotter phases are more metal enriched, both because the cooler phases make up most of the initially unenriched mass of the ISM, and the hotter phases are more directly populated by recent enrichment events, 3) the cooler, denser phases, particularly for supernova injected elements, are more well mixed than the hot phases of the ISM, 4) the PDFs of metal mass fraction are best fit by a log-normal + power law function, and therefore 5) the median metallicity available for star formation rates lies below the mean galactic value. In the case of supernova injected elements, enrichment proceeds quickly through the HIM over the entire galaxy. For AGB injected elements, enrichment proceeds through the WIM and WNM, leading to longer mixing timescales and larger metal fraction variations in the ISM.

@@

This toy model provides a physical intuition for the general trends in the PDFs presented here across ISM phases. The HIM is comprised predominantly of gas from individual enrichment events that, due to its low density, easily create large power-law tails beginning at very high metal fractions. Whatever ambient component of the HIM that exists is well mixed, leading to a narrow and sub-dominant log-normal component at low-metallicities. Cold, dense gas, which is almost never directly impacted by these individual enrichment events, is enriched almost entirely by diffusion, and thus has an almost completely log-normal PDF with very little, if any, power-law tail.

@@

%
% A.E. paragraph may need work
%
Mixing timescales are likely proportional to the eddy turnover times at the injection scale \citep{PanScannapieco2010, Colbrook2017} and the properties of turbulence in the ISM \citep{YangKrumholz2012}. Mixing within a phase is likely dependent upon the phase's sound speed and velocity dispersion. This would imply rapid mixing timescales in the WIM and HIM, with typical velocity dispersions of $\sim$~30 km~s$^{-1}$ and $\sim$~100~km~s$^{-1}$ respectively, and long mixing timescales in cold, dense gas ($\sim$ 1 km s$^{-1}$). Our results show, however, that in general the WIM and HIM are the least well mixed, while the CNM is the most well mixed across the galaxy. This is particularly curious as the sound crossing time of the HIM across the galaxy ($\sim 1$ kpc) is $\sim$10~Myr, compared to $\sim$~1~Gyr for the coldest gas. It must be that the HIM is far from equillibrium throughout the simulation, in part due to the continual enrichment by ongoing supernovae. Newly enriched gas must mix through the HIM on galaxy scales, becoming well mixed and less enriched by the time it cools into the CNM and eventually star forming gas. Since elements produced in AGB winds do not directly enter the HIM, mixing is less efficient driving larger variations across the galaxy. The idea of metals processing first through the hot ISM has been proposed before to explain observed metallicity trends in the outskirts of more massive galaxies \citep{Tassis2008,Werk2011}.

@@

Previous work on the evolution of metals in the ISM varies from studies concerning the advection of passive scalars in idealized turbulent boxes \citep[e.g.][]{Pope1991, PanScannapieco2010, PanScannapiecoScalo2012, PanScannapiecoScalo2013, YangKrumholz2012, SurPanScannapieco2014, Colbrook2017}
% cite passive scalar reviews by Warhaft 2000 in fluid mech and scalo + Elmegreen 2004?
to global galaxy models studying generalized advection and mixing of passive scalars \citep[e.g.][]{deAvillez2002,Petit2015,Goldbaum2016} to models with more detailed self-consistent metal enrichment \citep[e.g.][]{Revaz2009,Escala2018}.

@@

More detailed models of global galaxy evolution have generally focused on spatial correlations and mixing timescales of initially asymmetric fields, without concern for ongoing enrichment \citep[e.g.][]{deAvillez2002,Petit2015} or focused primarily on the evolution of stellar abundance patterns and metallicity distribution functions \citep[e.g][]{Jeon2017,Hirai2017,Escala2018}, without directly examining the evolution of gas-phase metallicity PDFs.

@@

%We discuss possible caveats to this in Section~\ref{sec:caveats}. Does this extend to more massive? how does DM potential affect this? Are general trends still true in much larger galaxies? or are things very different once HIM crossing times are more like 100 Myr (aka MW size galaxies).
